---
title: "Chapter 1: Introduction"
author: "Bodo Burger"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "figures/")
library(ElemStatLearn)
set.seed(3)
```

# Examples of learning problems

## Table 1.1. Spam data

*skipped for now*

## Figure 1.1. Scatterplot matrix of prostate cancer data

```{r figure-01-01-prostate,fig.asp=1}
df.prostate = ElemStatLearn::prostate

pairs(df.prostate[, c("lpsa", "lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45")],
      col = "blueviolet", cex = .5, cex.axis = .5)
```


## Figure 1.2. Examples of handwritten digits from U.S. postal envelopes

```{r Zip-data, message=FALSE, warning=FALSE, include=FALSE}
# The following code is taken from the help file of the zip.train dataset
# from the R package ElemStatLearn
findRows <- function(zip, n) {
  # Find  n (random) rows with zip representing 0,1,2,...,9
  res = vector(length = 10, mode = "list")
  names(res) = 0:9
  ind = zip[,1]
  for (j in 0:9) {
    res[[j + 1]] = sample(which(ind == j), n) 
  }
  return(res)
}

digits = vector(length = 10, mode = "list")
names(digits) = 0:9
rows = findRows(zip.train, 6)
for (j in 0:9) {
    digits[[j+1]] = do.call("cbind", lapply(as.list(rows[[j+1]]), 
                            function(x) zip2image(zip.train, x)))
}
im = do.call("rbind", digits)
```

```{r figure-01-02-zip, echo=FALSE}
image(im, col = gray(256:0/256), zlim = c(0,1), xaxt = 'n', yaxt = 'n')
```


## Figure 1.3. DNA microarray data

The data is from the NCI60 cancer microarray project (http://genome-www.stanford.edu/nci60).

```{r figure-01-03-nci}
df.nci = ElemStatLearn::nci
heatmap(df.nci[sample(nrow(df.nci), 100), ], Colv = NA, Rowv = NA, scale = "row",
        cexRow = .5, cexCol = .5, col = colorRampPalette(rev(c("green", "black", "red")))(15))
```

There are no grey cells representing missing values because these have already imputed and
the names of the genes (rows) are not available.
